---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r, echo=TRUE, include=FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

```{r, echo=TRUE}
activity <- read.csv('activity.csv')
activity$date <- as.Date(activity$date)
```


## What is mean total number of steps taken per day?

```{r, fig.height=10, fig.width=10}
activity2 <- aggregate(steps ~ date, data = activity, FUN = 'sum')
hist(activity2$steps,breaks = 53, xlab = 'Steps', main = 'Total Steps Each Day')
mean(activity2$steps)
median(activity2$steps)
```

### Mean and median total number of steps taken per day are 10766.19 and 10765 respectively


## What is the average daily activity pattern?

```{r, fig.height=10, fig.width=10}
#Plot of the 5-minute interval and the average number of steps taken
activity3 <- activity %>% group_by(interval) %>% summarize(steps=mean(steps, na.rm=TRUE))
ggplot(data = activity3, aes(x=interval, y=steps)) + geom_line()
head(arrange(activity3, desc(steps)))
```

### Interval 835 on average across all the days in the dataset, contains the maximum number of steps

## Imputing missing values

```{r}
#Calculate and report the total number of missing values in the dataset
length(which(is.na(activity$steps)))
```

### There are 2304 rows with missing values for steps

### Impute to replace NA with mean of steps of that interval

```{r}
missing <- filter(activity, is.na(activity$steps))
not_miss <- filter(activity, !is.na(activity$steps))
for(i in 1:nrow(activity3)){
  for(j in 1:nrow(missing))
    if(activity3$interval[i]==missing$interval[j]){
    missing$steps <- activity3$steps
  }
}
#combine and arrange the datasets
activity4 <- rbind(missing, not_miss)
activity5 <- arrange(activity4, date, interval)
activity6 <- aggregate(steps ~ date, data = activity5, FUN = 'sum')
```
### Dataset was divided into missing and non missing and then using for loop missing values were replaced by mean of steps in that interval

### Histogram of the total number of steps taken each day after imputation
```{r, fig.height=10, fig.width=10}
hist(activity6$steps,breaks = 53, xlab = 'Steps', main = 'Total Steps Each Day')
mean(activity6$steps)
median(activity6$steps)
```

### mean and median were 10766.19. No change in mean but a slight change by 1.19 was noticed in median


## Are there differences in activity patterns between weekdays and weekends?

### create new factor variable in the dataset with two levels -- "weekday" and "weekend" 
```{r}
activity5$tmp <- weekdays(activity5$date)
activity5$day <- ifelse(activity5$tmp=="Saturday", "weekend",
                 ifelse(activity5$tmp=="Sunday","weekend", "weekday"))
activity5$tmp <- NULL
activity5$day <- factor(activity5$day, levels = c("weekend", "weekday"))
activity7 <- activity5 %>% group_by(day, interval) %>% summarize(steps=mean(steps, na.rm=TRUE))
head(activity7)
```

### panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r, fig.height=10, fig.width=10}
ggplot(data = activity7, aes(x=interval, y=steps)) + geom_line() + 
  facet_wrap(~day, ncol=1, scales = "free_y") + theme_bw() +
  labs(x='Interval', y='Number of steps')

```

### Difference is activity pattern between weekdays and weekends were detected as shown in the plot above


